Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models

### 2024-09-05 manuscript.Rmd
updated abstract

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's `discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian \textit{generalized dynamic phylogenetic models} (GDPMs) that
overcomes these limitations and synthesizes the strengths of existing 
phylogenetic comparative methods into a single generalizable framework for 
dynamic phylogenetic causal inference. We formally and conceptually introduce
this class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
fitting GDPMs in Stan. We highlight previous studies that have used GDPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

>>>>

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of biological or cultural traits over evolutionary time. Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, Ornstein-Uhlenbeck models, and Pagel's `discrete' method. While these methods are appropriate under specific scenarios, they also limited by often unrealistic assumptions about the evolutionary process and limits on the  number and type of traits that researchers can include. Here, we develop a novel class of \textit{generalized dynamic phylogenetic models} (GDPMs) that overcomes these limitations and synthesizes the strengths of existing methods into a generalizable framework for dynamic phylogenetic inference. Treating the phylogeny as an implicit time series, GDPMs model the co-evolution of any number of traits undergoing both selection and drift, capable of inferring directed evolution (X -> Y vs Y -> X), feedback (X <-> Y), and contigencies (e.g., first X, then Y). We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan. We provide a short tutorial for fitting these models and present several empirical applications of GDPMs to make inferences about trait coevolution, including analyses of brain size in non-human primates and social complexity in human societies. ADD BRIEF SUMMARY/HIGHLIGHTS OF THE EMPIRICAL RESULTS.
\end{center}
